---
title: "Final Project: Part 1"
format: html
author: "Audrey Longan"
date: 3-17-26
execute: 
  echo: true
  warning: false
---

# Data Description

My **honors capstone data** will be used for this project. I collected this data from the spring of 2025 until the end of fall 2025. I also joined additional variables from the [EPA’s StreamCat dataset](https://www.epa.gov/national-aquatic-resource-surveys/streamcat-dataset) (Hill et al. 2016). Finally, mean runoff (mm), mean elevation (m), watershed area (km2), the percent base stream flow index, mean annual precipitation from 1991-2020 (mm), mean annual temperature from 1991-2020 (C°), and mean percent lithological calcium oxide (CaO) content within each watershed were downloaded from

- My collected data has 279 rows and 28 variables, 11 character variables, 15 double variables, and 2 time variables.

- **Methods for Data Collection and variables collected:**

  - Do a 50 m transect on both sides of the river, 1 meter out from shore, recorded plants every time seen, within 1 meters of each side of the tape (overall 4 meters recorded)

  - When a stream is less than or equal to 2 meters wide, I only surveyed one transect in the middle of the river, to cover the full two meters.

  - [Every time a plant was found I measured:]{.underline}

    - Depth (m)

    - Velocity (m/s)

    - Substrate size (mm)

    - Plant name and number of individuals or range/cover area on the transect

  - [Measure at each transect overall:]{.underline}

    - canopy cover facing north, south, east, and west, \[these numbers were averaged, then multiplied by 4 and divided by product of 17\*4 (freshwater ecology standard)\]

    - the percentage of the 50-meter transects that were riffle, pool, run, or glide.

    - water temperature (°C)

    - pH

    - conductivity(ms/cm)

    - TDS (g/L)

    - Dissolved oxygen (mg/L) and Dissolved oxygen (%)

    - Salinity (parts per thousand)

- **Research Questions:**

  - \(1\) How does abundance and diversity differ between site and transect? (2) Are there different environmental conditions that support specific species? (3) Which environmental factors lead to higher diversity? (4) Where do taxa tend to occur when they are not ubiquitous or rare?

## Data Cleaning

*reformat data cleaning section*

1.  Do you need to reformat any variables into different types (e.g. factors, time, dates, strings)? Or remove information from variable values?

> Yes, I changed the substrate and velocity variables to remove "\>" or "\<" symbols so that they can be numeric variables. I did this with mutate().

2.  Do you need to deal with any missing data, especially missing data coded other than NA?

> Yes, I removed NA data from the velocity variable and change NA values in the depth variable to zero. There is not any missing data that is not coded as NA. I did this with mutate().

3.  Do you need to filter your data? How?

> No, I don't need to do any filtering.

4.  Do you need to create any new variables? What variables? How?

> - using mutate()
>
> - a stem count variable called calc_stem_abundance_m2 by multiplying number of individuals by 0.09 (accounting for each individual as 900 cm\^2 on average for an estimate)
>
> - a final_abundance variable by using cover area when the value for number_individuals is NA or individuals are greater than or equal to 150 (this may be a less accurate estimate of individuals so using cover area is a better estimate in this case). I also added a clause for using the cover area variable instead of individuals for annual grass if individuals are greater than 5 (as it is difficult to count individual blades of grass)
>
> - Add a new variable of rooted_v_floating by combining the values of emergent and submergent into one value of "rooted"
>
> - creating a presence/absence column for each species at each site with 1 indicating presence

5.  Do you need to add new data (join) to your data? What data? How?

> Yes, I need to join additional environmental variables from the StreamCat dataset to my new abundance and presence/absence datasets. I will most likely use a left_join as I want to keep everything in the abundance and presence/absence datasets and only keep the data that matches from the StreamCat dataset. Greater understanding of how this will be accomplished will be known once I explore the StreamCat dataset further and determine exactly what needs to be done to join the datasets.

6.  Do you need pivot your data in any way? Why? How?

> Yes, I needed to pivot_wider to create the initial presence/absence matrix and then pivot_longer to add the environmental variables to the new presence/absence dataset.

7.  Do you need to summarize any of the variables? Which ones? How?

> Yes. I will be grouping my data by species, transect location and site. Then, I need to take the sum of the final_abundance, and calculate the mean for all my environmental variables, for each of the groupings.

8.  What other aspects of your data need to be “fixed” in order to make your data visualizations?

> I also used the janitor package to clean my variable names and make it easier to fully clean my data.

9.  Are there any variables you can exclude from your data?

> Yes, I excluded the variables: location_coordinates, date, observers, start_time, end_time, as they weren't important for my analysis.

**still need to comment out the rest of my code, I commented out the major important parts so far** **need to add labels to all my code chunks**

**need to add updated graph and copy this into the portfolio** **need to make correlogram better later**

**update answers to questions**

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(janitor)
library(vegan)
library(mosaic)
library(car)
library(ellipse)
library(RColorBrewer)
```

```{r}
#| label: reading-in-data
#| include: false
macrophytes <- read_csv("data-raw/macrophytes_full.csv")
macrophytes <- janitor::clean_names(macrophytes)
```

```{r}
#| label: making cover variable
#| eval: false

macrophytes |>
  mutate(calc_stem_abundance_m2 = number_individuals*0.09) |> #may change 0.09 to a better number to accurately represent size of plants, 30 cm^2 or 0.09 m^2, seems right
  mutate(final_abundance = case_when(is.na(number_individuals) ~ cover_area_m_2, 
                                     number_individuals >= 150 ~ cover_area_m_2,
                                     str_detect(aquatic_plant_species, "annual grass") & 
                                       !is.na(cover_area_m_2) & 
                                       number_individuals > 5 ~ cover_area_m_2,
                                     .default = calc_stem_abundance_m2
                                     )
  ) |>
  write_csv("data-raw/macrophytes_clean.csv")
# |> 
# select(number_individuals, calc_stem_abundance_m2, cover_area_m_2, aquatic_plant_species, total_abundance)
```

```{r}
#| eval: false
macrophytes_abun <- read_csv("data-raw/macrophytes_clean.csv", col_select = -c(location_coordinates, date, observers, start_time, end_time)) |>
  drop_na(velocity_m_s) |>
  mutate(substrate_size_mm = as.numeric(case_when(
          substrate_size_mm == "<2" ~ "1",
          substrate_size_mm == ">300" ~ "300",
          .default = substrate_size_mm
         )),
        velocity_m_s = as.numeric(case_when(
          velocity_m_s == "<0.1" ~ "0.05",
          .default = velocity_m_s
        )),
        depth_m = case_when(
          depth_m == "NA" ~ 0,
          .default = depth_m
        ),
        rooted_v_floating = case_when(
          emergent_submergent_or_floating == "emergent" ~ "rooted",
          emergent_submergent_or_floating == "submergent" ~ "rooted",
          emergent_submergent_or_floating == "partly emergent" ~ "rooted",
          emergent_submergent_or_floating == "floating" ~ "floating"
        ))
```

```{r}
#| label: creating-abundance-dataset
#| eval: false


# deciding what metric to use, mean or median for averaging all my numbers for each species, location, and bank
summary(macrophytes_abun)

macrophytes_abun |>
  count(substrate_size_mm)

macrophytes_abun |>
  count(velocity_m_s)

macrophytes_abun |>
  count(depth_m)

#creating total abundance for each species, bank, and site
macrophytes_abun |>
  group_by(aquatic_plant_species, transect_location_rb_lb, location_name, rooted_v_floating, water_body, season) |>
    summarise(Calc_total_abundance_m2 = sum(final_abundance), 
              canopy_cover_mean = mean(canopy_cover_0_to_1_ratio),
              temperature_c_mean = mean(temperature_c),
              p_h_mean = mean(p_h),
              conductivity_mean = mean(conductivity_m_s_cm),
              salinity_mean = mean(salinity),
              tds_g_l_mean = mean(tds_g_l),
              do_percent_mean = mean(do_percent),
              do_mg_l_mean = mean(do_mg_l),
              riffle_mean = mean(riffle),
              pool_mean = mean(pool),
              run_mean = mean(run),
              glide_mean = mean(glide),
              depth_m_mean = mean(depth_m),
              substrate_size_mm_mean = mean(substrate_size_mm),
              velocity_m_s_mean = mean(velocity_m_s),
              .groups = 'drop') |> 
 write_csv("data-clean/macrophytes_corr.csv")
```

```{r}
#| label: presence-absence-dataset-creation
#| eval: false

presence_absence_factors <- macrophytes_abun |>
    group_by(location_name, transect_location_rb_lb, water_body, season) |>
    summarise(canopy_cover_mean = mean(canopy_cover_0_to_1_ratio),
              temperature_c_mean = mean(temperature_c),
              p_h_mean = mean(p_h),
              conductivity_mean = mean(conductivity_m_s_cm),
              salinity_mean = mean(salinity),
              tds_g_l_mean = mean(tds_g_l),
              do_percent_mean = mean(do_percent),
              do_mg_l_mean = mean(do_mg_l),
              riffle_mean = mean(riffle),
              pool_mean = mean(pool),
              run_mean = mean(run),
              glide_mean = mean(glide),
              .groups = 'drop') |>
  write_csv("factorsfor_presabs_only.csv")

presence_absence_matrix <- macrophytes_abun |>
  select(location_name, transect_location_rb_lb, aquatic_plant_species) |> # Keep only unique occurrences
  distinct() |>
  mutate(Present = 1) |> # Add a column with value 1 for presence
  pivot_wider(
    id_cols = c(location_name, transect_location_rb_lb),
    names_from = aquatic_plant_species,
    values_from = Present,
    values_fill = 0 # Automatically fills non-observed combinations with 0
  ) |>
  pivot_longer(cols = -c(location_name, transect_location_rb_lb), names_to = "aquatic_plant_species", values_to = "presence_or_not") |>
  write_csv("data-clean/presence_absence_only.csv")

```

```{r}
#| label: abundance-diversity-indices
#| eval: false

macrophytes_final_original <- read_csv("data-clean/macrophytes_corr.csv")

new_variables <- read_csv("newvariables_wSites.csv")

macrophytes_final <- macrophytes_final_original |>
  full_join(new_variables, by = c("location_name" = "Location_Name")) #joining GIS variables to abundance data

macrophytes_final[16, 1] <- "Lunularia cruciata" #fixing two individual entries to be the correct names
macrophytes_final[17, 1] <- "Lunularia cruciata"

#between sites location only, not transect, transect has been included here
abundance_matrix <- macrophytes_final |>
  select(location_name, aquatic_plant_species, Calc_total_abundance_m2) |>
  group_by(aquatic_plant_species, location_name) |>
  summarise(Calc_total_abundance_m2 = sum(Calc_total_abundance_m2)) |>
  pivot_wider(names_from = aquatic_plant_species, values_from = Calc_total_abundance_m2, values_fill = 0) |>
  column_to_rownames("location_name")

# 1. Shannon Index
shannon <- diversity(abundance_matrix, index = "shannon")

# 2. Simpson Index (Gini-Simpson: 1 - D)
simpson <- diversity(abundance_matrix, index = "simpson")

# 3. Species Richness (S)
richness <- specnumber(abundance_matrix)

# 4. Pielou's Evenness (J)
evenness <- shannon / log(richness)

diversity_indices_results <- data.frame(
  Site = rownames(abundance_matrix),
  Shannon = shannon,
  Simpson = simpson,
  Richness = richness,
  Evenness = evenness
)

print(diversity_indices_results)

#overall indices
# 1. Create a single 'Overall' row by summing species abundances
overall_abundances <- abundance_matrix %>%
  summarise(across(everything(), sum))

# 2. Calculate indices for the Overall landscape
overall_results <- data.frame(
  Site = "OVERALL",
  Richness = specnumber(overall_abundances),
  Shannon = diversity(overall_abundances, index = "shannon"),
  Simpson = diversity(overall_abundances, index = "simpson"),
  Evenness = diversity(overall_abundances) / log(specnumber(overall_abundances))
)

# 3. Combine with your individual site results
final_table <- diversity_indices_results %>% # 'diversity_indices_results' is the table from the previous step
  bind_rows(overall_results)
```

#Visualization 1: Abundance of Different Species across Sites

```{r}
#| label: abundance-barchart
#| height: 5
#| width: 8

macrophytes_final |>
  select(location_name, aquatic_plant_species, Calc_total_abundance_m2) |>
  ggplot(aes(x = Calc_total_abundance_m2, fill = aquatic_plant_species, y = fct_relevel(location_name, "By Safeway Redo", "Creek Walk Start", "Schulte's Bridge (redo)", "Garland Ranch", "Lower Garzas (SLP)",  "Rosie's Bridge", "SLP Water Monitoring Site", "Tularcitos", "Cachagua Park"))) + 
  geom_col() + #makes the geometry of the plot bars
  labs(
    #creates the x and y axis labels
    x = expression("Total Abundance " *  (m^2)), 
    y = "Location",
    fill = "Aquatic Plant Species",
    alt = "Horizontal stacked bar chart showing total abundance in square meters of aquatic plant species across creek locations from upstream to downstream. Locations include CR (Carmel River) - Cachagua Park, Tularcitos Creek, San Clemente Creek, CR - Rosie's Bridge, Garzas Creek, CR - Garland Ranch, CR - Schulte's Bridge, Potrero Creek, and CR - By Safeway. CR - By Safeway has the highest total abundance, followed by Garzas Creek and CR - Schulte's Bridge. A color legend lists 16 aquatic plant species.") +
  theme_bw() + #This code got rid of the gray background 
  theme(legend.text = element_text(size = 12), axis.text = element_text(size = 12), axis.title = element_text(size = 16), legend.title = element_text(size = 16)) +
  scale_fill_viridis_d(option = "mako", begin = 0.25, end = 0.95) + #increasing text size for all of the aspects of the graph
  scale_y_discrete(labels = c("Tularcitos" = "Tularcitos Creek", "Schulte's Bridge (redo)" = "CR - Schulte's Bridge", "Garland Ranch" = "CR - Garland Ranch", "By Safeway Redo" = "CR - By Safeway", "Cachagua Park" = "CR - Cachagua Park", "Rosie's Bridge" = "CR - Rosie's Bridge", "Lower Garzas (SLP)" = "Garzas Creek", "SLP Water Monitoring Site" = "San Clemente Creek", "Creek Walk Start" = "Potrero Creek"))

  
```

**need to make this better later** 

#Visualization 2: Correlogram of important variables against each other

```{r}
#| label: correlogram-important predictors
#| height: 5
#| width: 8
 
cor_data <- cor(macrophytes_final[c(8:22, 29:34, 36)], use = "complete.obs") #specifying cor data
 
# Build a Pannel of 100 colors with Rcolor Brewer
my_colors <- brewer.pal(5, "Spectral")
my_colors <- colorRampPalette(my_colors)(100)
 
# Order the correlation matrix
ord <- order(cor_data[1, ])
cor_data_ord <- cor_data[ord, ord]
plotcorr(cor_data_ord , col=my_colors[cor_data_ord*50+50] , mar=c(1,1,1,1)  )
```
