---
title: "Project: STAT 210 Class Example"
format: html
author: "Audrey Longan"
date: 3-17-26
execute: 
  warning: false
  eval: true
---

# Data Description

My **honors capstone data** will be used for this project. I collected this data from the spring of 2025 until the end of fall 2025. I also joined additional variables from the [EPA’s StreamCat dataset](https://www.epa.gov/national-aquatic-resource-surveys/streamcat-dataset) (Hill et al. 2016). Finally, mean runoff (mm), mean elevation (m), watershed area (km2), the percent base stream flow index, mean annual precipitation from 1991-2020 (mm), mean annual temperature from 1991-2020 (C°), and mean percent lithological calcium oxide (CaO) content within each watershed were downloaded from

- My collected data has 279 rows and 28 variables, 11 character variables, 15 double variables, and 2 time variables.

- **Methods for Data Collection and variables collected:**

  - Do a 50 m transect on both sides of the river, 1 meter out from shore, recorded plants every time seen, within 1 meters of each side of the tape (overall 4 meters recorded)

  - When a stream is less than or equal to 2 meters wide, I only surveyed one transect in the middle of the river, to cover the full two meters.

  - [Every time a plant was found I measured:]{.underline}

    - Depth (m)

    - Velocity (m/s)

    - Substrate size (mm)

    - Plant name and number of individuals or range/cover area on the transect

  - [Measure at each transect overall:]{.underline}

    - canopy cover facing north, south, east, and west, \[these numbers were averaged, then multiplied by 4 and divided by product of 17\*4 (freshwater ecology standard)\]

    - the percentage of the 50-meter transects that were riffle, pool, run, or glide.

    - water temperature (°C)

    - pH

    - conductivity(ms/cm)

    - TDS (g/L)

    - Dissolved oxygen (mg/L) and Dissolved oxygen (%)

    - Salinity (parts per thousand)

- **Research Questions:**

  - \(1\) How does abundance and diversity differ between site and transect? (2) Are there different environmental conditions that support specific species? (3) Which environmental factors lead to higher diversity? (4) Where do taxa tend to occur when they are not ubiquitous or rare?

# Data Cleaning

*reformat data cleaning section*

1.  Do you need to reformat any variables into different types (e.g. factors, time, dates, strings)? Or remove information from variable values?

> Yes, I changed the substrate and velocity variables to remove "\>" or "\<" symbols so that they can be numeric variables. I did this with mutate().

2.  Do you need to deal with any missing data, especially missing data coded other than NA?

> Yes, I removed NA data from the velocity variable and change NA values in the depth variable to zero. There is not any missing data that is not coded as NA. I did this with mutate().

3.  Do you need to filter your data? How?

> No, I don't need to do any filtering.

4.  Do you need to create any new variables? What variables? How?

> - using mutate()
>
> - a stem count variable called calc_stem_abundance_m2 by multiplying number of individuals by 0.09 (accounting for each individual as 900 cm\^2 on average for an estimate)
>
> - a final_abundance variable by using cover area when the value for number_individuals is NA or individuals are greater than or equal to 150 (this may be a less accurate estimate of individuals so using cover area is a better estimate in this case). I also added a clause for using the cover area variable instead of individuals for annual grass if individuals are greater than 5 (as it is difficult to count individual blades of grass)
>
> - Add a new variable of rooted_v_floating by combining the values of emergent and submergent into one value of "rooted"
>
> - creating a presence/absence column for each species at each site with 1 indicating presence

5.  Do you need to add new data (join) to your data? What data? How?

> Yes, I need to join additional environmental variables from the StreamCat dataset to my new abundance and presence/absence datasets. I will most likely use a left_join as I want to keep everything in the abundance and presence/absence datasets and only keep the data that matches from the StreamCat dataset. Greater understanding of how this will be accomplished will be known once I explore the StreamCat dataset further and determine exactly what needs to be done to join the datasets.

6.  Do you need pivot your data in any way? Why? How?

> Yes, I needed to pivot_wider to create the initial presence/absence matrix and then pivot_longer to add the environmental variables to the new presence/absence dataset.

7.  Do you need to summarize any of the variables? Which ones? How?

> Yes. I will be grouping my data by species, transect location and site. Then, I need to take the sum of the final_abundance, and calculate the mean for all my environmental variables, for each of the groupings.

8.  What other aspects of your data need to be “fixed” in order to make your data visualizations?

> I also used the janitor package to clean my variable names and make it easier to fully clean my data.

9.  Are there any variables you can exclude from your data?

> Yes, I excluded the variables: location_coordinates, date, observers, start_time, end_time, as they weren't important for my analysis.

**need to make correlogram better later**

**commented out all code; need to add this to the portfolio**

**update answers to questions in data cleaning**

**add descriptions under visualizations, alt text, and captions**

```{r}
#| label: setup
#| include: false

#loading all of my packages
library(tidyverse)
library(janitor)
library(vegan)
library(mosaic)
library(car)
library(ellipse)
library(RColorBrewer)
library(viridis)
```

```{r}
#| label: reading-in-data
#| include: false


macrophytes <- read_csv("data-raw/macrophytes_full.csv") #reading in my original raw data before cleaning
macrophytes <- janitor::clean_names(macrophytes) #cleaning my variable names
```

```{r}
#| label: making-cover-variable
#| include: false

macrophytes |>
  mutate(calc_stem_abundance_m2 = number_individuals*0.09) |> #using 0.09 m^2 as an average cover area for each individual plant seen and creating a stem abundance variable with just that data
  mutate(final_abundance = case_when(is.na(number_individuals) ~ cover_area_m_2, 
                                     number_individuals >= 150 ~ cover_area_m_2,
                                     str_detect(aquatic_plant_species, "annual grass") & 
                                       !is.na(cover_area_m_2) & 
                                       number_individuals > 5 ~ cover_area_m_2,
                                     .default = calc_stem_abundance_m2
                                     ) #combining my stem abundance data with my regular cover area data with certain specifications for when I will use which variable for a certain species
  ) |>
  write_csv("data-raw/macrophytes_clean.csv") #writing this data into a new csv
```

```{r}
#| label: editing-environmental-variables
#| include: false

macrophytes_abun <- read_csv("data-raw/macrophytes_clean.csv", col_select = -c(location_coordinates, date, observers, start_time, end_time)) |> #reading in previously edited csv and removing unnecessary variables
  drop_na(velocity_m_s) |> #NAs removed from velocity
  mutate(substrate_size_mm = as.numeric(case_when(
          substrate_size_mm == "<2" ~ "1",
          substrate_size_mm == ">300" ~ "300",
          .default = substrate_size_mm
         )), #substrate size variable changed so there are no longer <2 or >300 values
        velocity_m_s = as.numeric(case_when(
          velocity_m_s == "<0.1" ~ "0.05",
          .default = velocity_m_s
        )), #velocity variable changed so there are no longer <0.1 values
        depth_m = case_when(
          depth_m == "NA" ~ 0,
          .default = depth_m
        ), #NA values in depth changed to zero, this is a logical step made for my research
        rooted_v_floating = case_when(
          emergent_submergent_or_floating == "emergent" ~ "rooted",
          emergent_submergent_or_floating == "submergent" ~ "rooted",
          emergent_submergent_or_floating == "partly emergent" ~ "rooted",
          emergent_submergent_or_floating == "floating" ~ "floating"
        )) #new rooted versus floating variable created from the emergent-submergent-floating variable
```

```{r}
#| label: creating-abundance-dataset
#| include: false

#creating total abundance for each species, bank, and site
macrophytes_abun |>
  group_by(aquatic_plant_species, transect_location_rb_lb, location_name, rooted_v_floating, water_body, season) |> #grouping my data by species, transect location, and site; the rooted_v_floating, water_body, and season variables were added so they could be in the table but they do not change how the data is grouped based on the first 3 variables
    summarise(Calc_total_abundance_m2 = sum(final_abundance), 
              canopy_cover_mean = mean(canopy_cover_0_to_1_ratio),
              temperature_c_mean = mean(temperature_c),
              p_h_mean = mean(p_h),
              conductivity_mean = mean(conductivity_m_s_cm),
              salinity_mean = mean(salinity),
              tds_g_l_mean = mean(tds_g_l),
              do_percent_mean = mean(do_percent),
              do_mg_l_mean = mean(do_mg_l),
              riffle_mean = mean(riffle),
              pool_mean = mean(pool),
              run_mean = mean(run),
              glide_mean = mean(glide),
              depth_m_mean = mean(depth_m),
              substrate_size_mm_mean = mean(substrate_size_mm),
              velocity_m_s_mean = mean(velocity_m_s),
              .groups = 'drop') |> #means calculated for my variables
 write_csv("data-clean/macrophytes_corr.csv") #final dataset has been created for abundance
```

```{r}
#| label: presence-absence-dataset-creation
#| include: false

#Creating a presence-absence version of my dataset for each transect and site

presence_absence_factors <- macrophytes_abun |>
    group_by(location_name, transect_location_rb_lb, water_body, season) |> #grouping by site, transect, water body and season
    summarise(canopy_cover_mean = mean(canopy_cover_0_to_1_ratio),
              temperature_c_mean = mean(temperature_c),
              p_h_mean = mean(p_h),
              conductivity_mean = mean(conductivity_m_s_cm),
              salinity_mean = mean(salinity),
              tds_g_l_mean = mean(tds_g_l),
              do_percent_mean = mean(do_percent),
              do_mg_l_mean = mean(do_mg_l),
              riffle_mean = mean(riffle),
              pool_mean = mean(pool),
              run_mean = mean(run),
              glide_mean = mean(glide),
              .groups = 'drop') |> #means of my overall variables for each site were calculated
  write_csv("data-raw/factorsfor_presabs_only.csv") #presence-absence predictors dataset created


presence_absence_matrix <- macrophytes_abun |>
  select(location_name, transect_location_rb_lb, aquatic_plant_species) |> # Keep only unique occurrences
  distinct() |>
  mutate(Present = 1) |> # Add a column with value 1 for presence
  pivot_wider(
    id_cols = c(location_name, transect_location_rb_lb),
    names_from = aquatic_plant_species,
    values_from = Present,
    values_fill = 0 # Automatically fills non-observed combinations with 0
  ) |>
  pivot_longer(cols = -c(location_name, transect_location_rb_lb), names_to = "aquatic_plant_species", values_to = "presence_or_not") |>
  write_csv("data-raw/presence_absence_only.csv") #presence-absence variable matrix created and stored; the predictors dataset and presence-absence variable matrix were joined together outside of R to create one presence-absence dataset

```

```{r}
#| label: abundance-diversity-indices
#| include: false
#| message: false

macrophytes_final_original <- read_csv("data-clean/macrophytes_corr.csv") #reading in the original final dataset

new_variables <- read_csv("newvariables_wSites.csv") #reading in the dataset from StreamCat with the additional variables

macrophytes_final <- macrophytes_final_original |>
  full_join(new_variables, by = c("location_name" = "Location_Name")) #joining GIS StreamCat variables to abundance data

macrophytes_final[16, 1] <- "Lunularia cruciata" #fixing two individual entries to be the correct names as it previously only said "Lunularia spp., which was wrong"
macrophytes_final[17, 1] <- "Lunularia cruciata"

#Diversity Indices being calculated
abundance_matrix <- macrophytes_final |> 
  select(location_name, aquatic_plant_species, Calc_total_abundance_m2) |> #variables needed for abundance selected
  group_by(aquatic_plant_species, location_name) |> #data grouped by species and location
  summarise(Calc_total_abundance_m2 = sum(Calc_total_abundance_m2)) |> #abundance was summarized based on species and location from the group_by
  pivot_wider(names_from = aquatic_plant_species, values_from = Calc_total_abundance_m2, values_fill = 0) |> #this new summarised information is pivoted to a wide format for the diversity calculations below
  column_to_rownames("location_name")

# 1. Shannon-Weiner Diversity Index
shannon <- diversity(abundance_matrix, index = "shannon") 

# 2. Simpson's Diversity Index (Gini-Simpson: 1 - D)
simpson <- diversity(abundance_matrix, index = "simpson") 

# 3. Species Richness (S)
richness <- specnumber(abundance_matrix)

# 4. Pielou's Evenness (J)
evenness <- shannon / log(richness)

diversity_indices_results <- data.frame(
  Site = rownames(abundance_matrix),
  Shannon = shannon,
  Simpson = simpson,
  Richness = richness,
  Evenness = evenness
) #results of my diversity indices have been combined into a table

print(diversity_indices_results)

#overall indices
# 1. Create a single 'Overall' row by summing species abundances
overall_abundances <- abundance_matrix %>%
  summarise(across(everything(), sum)) 

# 2. Calculate indices for the Overall landscape
overall_results <- data.frame(
  Site = "OVERALL",
  Richness = specnumber(overall_abundances),
  Shannon = diversity(overall_abundances, index = "shannon"),
  Simpson = diversity(overall_abundances, index = "simpson"),
  Evenness = diversity(overall_abundances) / log(specnumber(overall_abundances))
) #summarizing the overall index into a table

# 3. Combine with your individual site results
final_table <- diversity_indices_results %>% # 'diversity_indices_results' is the table from the previous step
  bind_rows(overall_results)
#these results were copied into a table created outside of R
```

```{r}
#| label: variable-removal
#| include: false

#removal of variables from the dataset which were later determined to be redundant or unnecessary for graphics
macrophytes_final <- macrophytes_final |>
  select(-c(salinity_mean, tds_g_l_mean, do_percent_mean, glide_mean, StreamCat_data_csv_tmean9120ws, StreamCat_data_csv_catareasqkmrp100, StreamCat_data_csv_wsareasqkmrp100, StreamCat_data_csv_catareasqkm, watercress_inCatchments_2_AreaSqKM, watercress_inCatchments_2_LengthKM, watercress_inCatchments_2_Shape_Length_1))
  
```

# Visualization 1: Abundance of Different Species across Sites

```{r}
#| label: abundance-barchart
#| echo: false
#| fig-cap: "This is a caption for the first visualization"
#| fig-alt: "Horizontal stacked bar chart showing total abundance in square meters of aquatic plant species across creek locations from upstream to downstream. Locations include CR (Carmel River) - Cachagua Park, Tularcitos Creek, San Clemente Creek, CR - Rosie's Bridge, Garzas Creek, CR - Garland Ranch, CR - Schulte's Bridge, Potrero Creek, and CR - By Safeway. CR - By Safeway has the highest total abundance, followed by Garzas Creek and CR - Schulte's Bridge. A color legend lists 16 aquatic plant species."
#| fig-height: 5
#| fig-width: 8

macrophytes_final |>
  select(location_name, aquatic_plant_species, Calc_total_abundance_m2) |> #site, species, and abundance variables selected
  ggplot(aes(x = Calc_total_abundance_m2, fill = aquatic_plant_species, y = fct_relevel(location_name, "By Safeway Redo", "Creek Walk Start", "Schulte's Bridge (redo)", "Garland Ranch", "Lower Garzas (SLP)",  "Rosie's Bridge", "SLP Water Monitoring Site", "Tularcitos", "Cachagua Park"))) +  #ggplot created with x mapping to abundance, fill mapping to species, and y mapping to site, using fct_relevel to order my sites as upstream to downstream
  geom_col() + #makes the geometry of the plot bars
  labs(
    #creates the x and y axis labels as well as my legend title
    x = expression("Total Abundance " *  (m^2)), 
    y = "Location",
    fill = "Aquatic Plant Species") +
  theme_bw() + #This code got rid of the gray background 
  theme(legend.text = element_text(size = 12), axis.text = element_text(size = 12), axis.title = element_text(size = 16), legend.title = element_text(size = 16)) + #increasing text size for all of the aspects of the graph
  scale_fill_viridis_d(option = "mako", begin = 0.25, end = 0.95) + #using the viridis mako option as I like this color pallete the best
  scale_y_discrete(labels = c("Tularcitos" = "Tularcitos Creek", "Schulte's Bridge (redo)" = "CR - Schulte's Bridge", "Garland Ranch" = "CR - Garland Ranch", "By Safeway Redo" = "CR - By Safeway", "Cachagua Park" = "CR - Cachagua Park", "Rosie's Bridge" = "CR - Rosie's Bridge", "Lower Garzas (SLP)" = "Garzas Creek", "SLP Water Monitoring Site" = "San Clemente Creek", "Creek Walk Start" = "Potrero Creek")) #changing the names of my sites too look nicer on the graph

  
```

**need to make this better later**

# Visualization 2: Correlogram of important variables against each other

```{r}
#| label: correlogram-important predictors
#| echo: false
#| fig-cap: "This is a caption for the first visualization"
#| fig-alt: "add alt text"
#| fig-height: 5
#| fig-width: 8
 
cor_data <- cor(macrophytes_final[c(8:18, 22:27)], use = "complete.obs") #specifying correlogram data

# Renaming variables before plotting
new_labels <- c(
  "canopy_cover_mean" = "Canopy Cover",
  "temperature_c_mean" = "Temperature",
  "p_h_mean" = "pH",
  "conductivity_mean" = "Conductivity",
  "do_mg_l_mean" = "DO",
  "riffle_mean" = "Riffle (%)",
  "pool_mean" = "Pool (%)",
  "run_mean" = "Run (%)",
  "depth_m_mean" = "Depth",
  "substrate_size_mm_mean" = "Substrate Size",
  "velocity_m_s_mean" = "Velocity",
  "StreamCat_data_csv_bfiws" = "Base Stream Flow Index",
  "StreamCat_data_csv_elevws" = "Elevation",
  "StreamCat_data_csv_caows" = "CaO Content",
  "StreamCat_data_csv_runoffws" = "Runoff",
  "StreamCat_data_csv_precip9120ws" = "Precipitation",
  "StreamCat_data_csv_wsareasqkm" = "Watershed Area"
)

#establishing the new labels for the correlogram data
rownames(cor_data) <- new_labels[rownames(cor_data)]
colnames(cor_data) <- new_labels[colnames(cor_data)]
 
# Building a pannel of colors using the viridis mako option
my_colors <- viridis(100, option = "mako", begin = 0.25, end = 0.95)
 
# Orders the correlation matrix
ord <- order(cor_data[1, ])
cor_data_ord <- cor_data[ord, ord]

cor_data_ord |>
  plotcorr(col=my_colors[cor_data_ord*50+50] , mar=c(1,1,1,1)) #plots the correlogram
```

DO % as well as the glide variable were removed due to redundancy in other variables as well as limited data. Salinity and TDS were also removed as they are derivative calculations of conductivity which was kept. Air temperature was removed as well since it doesn't change much over the years and water temperature is a better metric for aquatic plants. **add to data cleaning section**
